---
title: "Exploratory analysis of RNAseq data"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    highlight: pygments
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, rows.print = 10)
```

[back to lesson's homepage](https://tavareshugo.github.io/data-carpentry-rnaseq/)

# Lesson objectives

* Understand the dataset being used
* Introducing a `matrix` object and how to convert it to a `tibble`


# Summary of dataset

In this lesson, we will apply some of the skills that we've gained so far to manipulate 
and explore a dataset from an RNAseq experiment. 

This lesson uses data from an experiment included in the 
[`fission` R/Bioconductor package](https://bioconductor.org/packages/release/data/experiment/vignettes/fission/inst/doc/fission.html). 
Very briefly, we have transcriptome data for:

* Two yeast strains: wild type ("wt") and _atf21del_ mutant ("mut")
* Each has 6 time points of osmotic stress time (0, 15, 30, 60, 120 and 180 mins)
* Three replicates for each strain at each time point

Let's say that you did this experiment yourself, and that a bioinformatician 
analysed it and provided you with four things:

* Information about each sample
* Read counts for all genes (a measure of the genes' expression) simply normalised to 
the size of each library (to account for the fact that different samples have more 
or less total number of reads)
* Normalised read counts for all genes but on a log scale and transformed to correct 
for a dependency between the mean and variance, which is typical in such count data
* Results from a statistical test that assessed the likelihood of your data 
assuming no difference between the first each of the other time points in WT cells


# Getting started

The data are provided within an `RData` file, which you can download and read 
into your R session.

First, let's clean our workspace (remove any objects created before), 
and also load the `tidyverse` package. 

It's also a good idea to check if your _working directory_ is correctly set to 
your workshop folder, which should contain a `data` folder within it (you can 
use `getwd()` to check your _working directory_).

```{r, eval = FALSE}
# Clean your workspace by removing any objects created beforehand
rm(list = ls())

# Load the tidyverse package
library(tidyverse)

# Download the data provided by your collaborator
dir.create("data")
download.file("https://github.com/tavareshugo/data-carpentry-rnaseq/blob/master/data/fission_data.RData?raw=true",
              destfile = "data/fission_data.RData",
              mode = "wb")

# Load the data
load("data/fission_data.RData")
```

```{r, echo = FALSE, message = FALSE}
# In the repository data is read from the current directory
# Load the tidyverse package
library(tidyverse)

# Load the data
load("data/fission_data.RData")
```

----

**Exercise:**

> Familiarise yourself with these data:
>
> - what kind of objects were you provided with? (hint: `class()`)
> - How many samples did you have? Is the design balanced?
> - How many genes do you have gene expression levels for?

----



### Matrix object

You might have noticed that `trans_cts` is a `matrix` object. We haven't found 
these before, and to produce visualisations of our data, it is convenient to 
start by converting our matrix of gene expression to a `tibble` (`data.frame`).

Matrices are a bit similar to `data.frame`, but they only contain values of a 
single type, in this case numeric values (whereas in a `data.frame` different 
columns can contain different types of data). 

```{r}
# Look at the first 10 rows and first 5 samples of trans_cts
trans_cts[1:10, 1:5]
```

You will also notice that gene names are contained in the row names of this 
matrix (they are not a column of their own!). 

To convert this matrix into a `tibble` object we can use the function `as_tibble()`:

```{r}
# Convert matrix to tibble
as_tibble(trans_cts)
```

But now we've lost our gene names! 
If we look at the function's help (`?as_tibble`), we can see that there's a way 
to solve this problem:

```{r}
# Convert matrix to tibble - add colnames to a new column called "gene"
trans_cts_tbl <- as_tibble(trans_cts, rownames = "gene")
trans_cts_tbl
```

----

**Exercise:**

* Convert the `norm_cts` matrix to a tibble and save it into a new object called `norm_cts_tbl`

----


----

[back to lesson's homepage](https://tavareshugo.github.io/data-carpentry-rnaseq/)
